Constraint Grammar as a SAT problem

### 2015-03-30 paper/nodalida2015.tex
small changes here and there

The ambiguous passage can be either a noun phrase, \emph{la\texttt{:det} casa\texttt{:n}} 
`the house'  or a verb phrase \emph{la\texttt{:prn} casa\texttt{:v pri p3}} `(he/she) marries her'. 
We add the following rules:

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of ``last'' is not applicable.
With these constraints, we get two solutions: 
\begin{enumerate}
\item []
\begin{verbatim}
"<la>"
        "el" det def f sg
"<casa>"
        "casa" n f sg
\end{verbatim}
\item [and]
\begin{verbatim}
"<la>"
        "lo" prn p3 f sg
"<casa>"
        "casar" v pri p3 sg
        "casar" v imp p2 sg
\end{verbatim}
\end{enumerate} 
% \texttt{det, n} and \texttt{prn, v pri p3, v imp p2}. 
The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} 
for free, and ordering of the rules doesn't matter. 
%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

The most important differences between the traditional and the SAT-based approach are summarised below:

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the word which is tagged as \texttt{prn} or
\texttt{det}. The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition \texttt{(1 n)} or \texttt{(1 v)}.
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions can be written as many
rules, each condition taking its turn to be in position to be removed
or selected. \\

@@

\noindent These three features might change the way that rules are written. 
We predict that less rules are needed; whether this holds in
the order of thousands of rules remains to be tested.
Getting rid of ordering could ease the task of the grammar writer,
removing the responsibility to estimate the best sequence of rule applications.
However, the implementation should not be too different from the
existing ones---our preliminary evaluation against VISL CG-3 shows
promising results in this regard.
\pagebreak

@@

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is just a first proof-of-concept implementation with no optimisations.

@@

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all ambiguities, and ran both VISL CG-3 and our implementation with the same grammar. 
Treating the chosen tagset as the gold standard, the disambiguation by
SATCG achieves 75.09 \% correct result and VISL CG-3 75.80 \%.
Similar patterns were observed with small (\textless{}20) rule sets written by
the authors; depending on the subset, SATCG and VISL CG-3 had
a difference of at most $\Mypm$ 1.5 \%. 
Introducing rules one by one up to 19, the
performance improved in a very similar rate, with less than 0.5 \%
difference between the systems at each new rule.

>>>>

The ambiguous passage can be either a noun phrase, \emph{la}\texttt{<det>} \emph{casa}\texttt{<n>} 
`the house'  or a verb phrase \texttt{la<prn>}  \emph{casa}\texttt{<v><pri><p3>} `(he/she) marries her'. 
We add the following rules:

@@

SAT solver performs a search, 
and starts building possible models that satisfy both constraints. 
In addition to the given constraints, we have default rules to emulate
the CG principles: an analysis is true if no rule affects it,
and at least one analysis for each word is true---the notion of
``last'' is not applicable.

With these constraints, we get two solutions. The interaction of the rules regarding \emph{la}  disambiguates the POS of \emph{casa} for free, and ordering of the rules doesn't matter.

\begin{enumerate}
\item [\texttt{1)}]
\begin{verbatim}
"<la>"
        "el" det def f sg
"<casa>"
        "casa" n f sg
\end{verbatim}
\item [\texttt{2)}]
\begin{verbatim}
"<la>"
        "lo" prn p3 f sg
"<casa>"
        "casar" v pri p3 sg
        "casar" v imp p2 sg
\end{verbatim}
\end{enumerate}

%Without more context or additional rules we cannot fully disambiguate the passage,
%but unlike with standard CG, we get only legal combinations in one solution.

@@

\noindent The most important differences between the traditional and the SAT-based approach are summarised below:

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the word which is tagged as \texttt{prn} or
\texttt{det}. The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{1 n} or \texttt{1 v}).
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions can be written as many
rules, each condition taking its turn to be in position to be removed
or selected. \\

@@

\noindent These three features might change the way that rules are written. 
We predict that less rules are needed; whether this holds in
the order of thousands of rules remains to be tested.
Getting rid of ordering could ease the task of the grammar writer,
since it removes the need to estimate the best sequence of rule
applications.

However, the implementation should not be too different from the
existing ones---our preliminary evaluation against VISL CG-3 shows
promising results in this regard.
\pagebreak

@@

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is currently just a naive proof-of-concept implementation with no optimisations.

@@

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all
ambiguities, and ran both SATCG and VISL CG-3 with the same grammar. 
Treating the corpus as the gold standard, the disambiguation by
SATCG achieves 75.09 \% correct result and VISL CG-3 75.80 \%.
Similar patterns were observed with small (\textless{}20) rule sets
written by the authors; depending on the subset, SATCG and VISL CG-3 had
a difference of at most $\Mypm$ 1.5 \%. 
Introducing rules one by one up to 19, the
performance improved in a very similar rate, with less than 0.5 \%
difference between the systems at each new rule.
